A Theory for Data Migration of Generated Evolving Information Systems

### 2022-10-30 2022 Migration/articleMigration.tex
Result after 30 Oct 2022 discussion

   Triples serve to represent data.
   A dataset $\dataset$ is a set of triples.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
   Each triple relates two atoms to a relation.
   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
   the ``thing'' that \atom{Peter} refers to has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.
   In this paper, the informal meaning is interesting only for the sake of illustration.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\dataset$
   in those cases where $\dataset$ is either clear from the context or irrelevant for the argument.
   A relation $r$ can serve as a container of pairs,
   as defined by the funciton $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\dataset\}
\end{equation}

@@

   A schema $\schema$ is a triple $\triple{\concepts_{\schema}}{\rels_{\schema}}{\rules_{\schema}}$,
   in which $\concepts_{\schema}$ is a finite set of concepts,
   $\rels_{\schema}$ is a finite set of relations,
   and $\rules_{\schema}$ is a finite set of rules.
   We will simply write $\concepts$, $\rels$, or $\rules$ if the schema is clear from the context or irrelevant for the argument.
   A rule in a schema serves to constrain the dataset,
   to ensure the semantic integrity of the dataset.
   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\id{sat}_u\{\Dataset\}$
   to determine whether rule $u$ is satisfied in dataset $\dataset$.
   The Ampersand compiler also generates code for a function $\id{viol}_u : \Dataset\rightarrow\id{Text}$
   to produce meaningful error messages in case rule $u$ is not satisfied.

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to its users.
   This section introduces the notion of information system as a quintuple $\la\dataset_{\infsys},\schema_{\infsys},\inst_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
   As before, we will omit the suffixes if $\infsys$ is obvious (i.e. clear from the context or irrelevant for the argument).
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\inst,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ and $\schema$ are defined as in sections~\ref{sct:Schemas} and~\ref{sct:Datasets};
   \item $\roles$ is a set of roles;
   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
   \item requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} are satisfied.
\end{itemize}
\end{definition}
   Let us discuss these items one by one.

   The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
   This means that every atom is an instance of a concept:
\begin{eqnarray}
   \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
\end{eqnarray}
   To do type checking based on the schema only,
   all relations in triples from the dataset must be ``known'' in the schema:
\begin{equation}
   \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels
\end{equation}

   In a typed information system, atoms in a relation have a type as declared in a relation.
   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
\begin{eqnarray}
   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\\
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
\end{eqnarray}

   To compare atoms for equality,
   the set $\rels$ must contain an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
   It must satisfy:
\begin{equation}
   \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
\end{equation}
   So, $\id{pop}_{\ident{C}}$ is defined by:
\begin{eqnarray}
   \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}| a\ \inst\ C\}
\end{eqnarray}

   Ampersand supports a special kind of rule that is used for specialization.
   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
\begin{equation}
   \label{eqn:specialization}
   \id{sat}_{A\isa B}\{\dataset\}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
\end{equation}
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.

   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
   Specialization rules are special for two reasons:
   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
   This allows for certain run-time optimizations, which are beyond the scope of this paper.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write specialization rules that make these comparisons possible.
   The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.

@@

   Every rule is assigned to a role.
   That role must ensure to satisfy that rule, after some event has violated the rule by changing the set of triples.
   In other words: If anyone edits in the dataset and violates a rule, someone or something has to fix the dataset to satisfy that rule again.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
   Sending an acknowledgement message will cause the rule to be satisfied again.

   Users have their own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use serves a purpose which we loosely call ``the business''.
   As a consequence, the data in a system changes continually.
   To preserve meaning, users are trying to maintain semantic constraints on the data,
   amidst of all changes that are going on around them.
   So, let us first define the constituent notions role and rule before defining information systems proper.

   A \define{role} is a name that identifies a group of users.
   The purpose of a role is to mention an individual user without knowing who that user is.
   In the sequel, when we talk about a role we actually mean a user (an arbitrary one) who fulfills that role in the system.

   The relation $\maintain$ links rules to roles.
   $o\ \maintain\ u$ means that role $o$ must keep the set $\viol{u}{\dataset}$ empty.
   If a rule $u$ has violations, i.e. $\viol{u}{\dataset}$ is not empty, there is a role whose job it is to remove those violations.
   The relation $\maintain$ being surjective means that all users together are keeping the violation set empty.

\subsection{Example}
\label{old IS}
   Having defined an information system in mathematical terms, let us give an example.
   For this purpose we use a subset of the language Ampersand
   because it makes the examples more appealing to read.
   Let us first define a dataset of seven triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
   This introduces a relation with name \verb#takes#,
   source concept \verb#Student#, and
   target concept \verb#Course#.
   It also introduces three triples:
\[\begin{array}{l}
   \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
   \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
   \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
\end{array}\]
   The informal meaning of this relation is that it states which students are taking which courses.
   So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor [Student*Module]
\end{verbatim}
   Summarizing, here we have a set of atoms, which contains \verb-"Peter"-, \verb-"Finance"-, etc.,
   a set of concepts $\{$\verb-Student-, \verb-Module-, \verb-Course-$\}$,
   an instance relation that states which atoms belong to which concept,
   a specialization relation, which is empty,
   a set of three relations, and
   a set of eight triples.
   The example in Ampersand defines precisely that dataset.

   Now let us define an information system.
   We define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
   Its constraint, $p_{\tt Enroll}$ is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to maintain it:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The Ampersand compiler derives the signature of every rule.
   This rule has signature $\pair{\text{\tt Student}}{\text{\tt Module}}$.
   The Ampersand compiler also derives the violation set as:
\[\begin{array}{l}
   \viol{\text{\tt EnrollRule}}{\dataset}\\
   \hspace{1cm}=\{\pair{s}{m}|s\ \text{\tt isEnrolledFor}\ m\ \wedge\not\exists c: s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{array}\]
   Notice that the current population yields an empty set of violations,
   so the rule is satisfied.
   Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
   we have now defined a small system.

@@

   We use two datasets: $\dataset$ and $\dataset'$.
   Before doing so, let us first define the disjoint union of two systems.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
      X\uplus Y&=&\{(x,0)|\ x\in X\}\ \cup\ \{(y,1)|\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))|\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))|\ (y_1,y_2)\in Y\}\end{array}\\
      X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
      X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
\end{eqnarray}
\end{definition}
Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.

@@

\begin{definition}[migration dataset]
   \begin{eqnarray}
      &&\begin{array}[t]{@{}l}
         \{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt\ :=\ }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
            \mid\ \begin{array}[t]{@{}l}
               \declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
               \{\pair{A}{A'},\pair{B}{B'}\}\subseteq\isa\cup\isa'\cup\flip{\isa}\cup\flip{\isa'}\}
               \end{array}
           \end{array}\\
           &\cup&\begin{array}[t]{@{}l}
            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)|\ C\in\concepts,C'\in\concepts',C=C'\}\\
           \end{array}
   \end{eqnarray}
\end{definition}

@@

\subsection{Changes}
% The purpose of this section is to explain why a dataset is structured the way it is.
   In the migration of a dataset we deal with changes to the elements that are not data:
   $\inst$, $\concepts$, and $\rels$.
   Such changes have further reaching consequences, however.
   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
   To satisfy definition~\ref{def:dataset}, $\dataset$ and $\atoms$ must change too,
   but the migration should try to preserve the maximal amount of data.

   This paper studies those changes and seeks to preserve the rules in definition~\ref{def:dataset}
   in such a way that large portions of migration can be automated.

>>>>

   Triples serve to represent data.
   A dataset $\dataset$ is a set of triples.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   This makes our theory valid for any kind of database that triples can represent,
   such as SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases
   Each triple relates two atoms to a relation.
   A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean in practice that
   the ``thing'' that \atom{Peter} refers to has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.
   In this paper, the informal meaning is interesting only for the sake of illustration.
   To save writing in the sequel, we will write $a\ r\ b$ to denote that $\triple{a}{r}{b}\in\dataset$
   in those cases where $\dataset$ is either clear from the context or irrelevant for the argument.
   A relation $r$ can serve as a container of pairs,
   as defined by the funciton $\id{pop}_r:\Dataset\rightarrow\Pair{\Atoms}{\Atoms}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\dataset\}
\end{equation}

@@

   A schema $\schema$ is a triple $\triple{\concepts_{\schema}}{\rels_{\schema}}{\rules_{\schema}}$,
   in which $\concepts_{\schema}$ is a finite set of concepts,
   $\rels_{\schema}$ is a finite set of relations,
   and $\rules_{\schema}$ is a finite set of rules.
   We will simply write $\concepts$, $\rels$, or $\rules$ if the schema is clear from the context or irrelevant for the argument.

   In a typed information system, atoms in a relation have a type as declared in a relation.
   For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
   We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
   We require that the source and target concepts are in the set of concepts:
\begin{eqnarray}
   \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts
\end{eqnarray}
   A rule in a schema serves to constrain the dataset,
   to ensure the semantic integrity of the dataset.
   For every rule $u$ in the schema, the Ampersand compiler generates code for the predicate $\sat{u}{\Dataset}$
   to determine whether rule $u$ is satisfied in dataset $\dataset$.
   The Ampersand compiler also generates code for a function $\id{viol}_u : \Dataset\rightarrow\id{Text}$
   to produce meaningful error messages in case rule $u$ is not satisfied.

\subsection{Information Systems}
\label{sct:Information Systems}
   The purpose of an information system is to make data meaningful to its users.
   Users have their own tasks and responsibilities
   and may work from different locations and on different moments.
   This collective use serves a purpose which we loosely call ``the business''.
   As a consequence, the data in a system changes continually.
   To preserve meaning, users are trying to maintain semantic constraints on the data,
   amidst of all changes that are going on around them.

   This section introduces the notion of information system as a quintuple $\la\dataset_{\infsys},\schema_{\infsys},\inst_{\infsys},\roles_{\infsys},\maintain_{\infsys}\ra$.
   As before, we will omit the suffixes if $\infsys$ is obvious (i.e. clear from the context or irrelevant for the argument).
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\inst,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema$ is defined as in sections~\ref{sct:Schemas};
   \item $\inst : \Atoms\times\Concepts$ is the instance relation between atoms from $\dataset$ and concepts from $\schema$;
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$; and
   \item requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} are satisfied.
\end{itemize}
\end{definition}
   Let us discuss these items one by one.

   The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
   This means that every atom is an instance of a concept:
   To do type checking based on the schema only,
   all concepts in the relation $\id{inst}$ and all relations in triples from the dataset must be ``known'' in the schema:
\begin{eqnarray}
   a\ \inst\ C&\Rightarrow&C\in\concepts\label{eqn:inst}\\
   \triple{a}{r}{b}\in\dataset&\Rightarrow&r\in\rels\label{eqn:define R}
\end{eqnarray}

\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
\end{eqnarray}

   Ampersand supports a special kind of rule that is used for specialization.
   The rule $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
\begin{equation}
   \label{eqn:specialization}
   \sat{A\isa B}{\dataset}\ \Leftrightarrow\ \forall a: a\ \inst\ A\rightarrow a\ \inst\ B
\end{equation}
   We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
   Specialization is needed to allow statements such as: ``An employee is a person'' or ``A human is a mammal''.
   All specialization rules in a schema together form a partial order $\isa : \Pair{\concepts}{\concepts}$.
   Specialization rules are special for two reasons:
   Firstly, the system keeps $A\isa B$ satisfied automatically by changing $\inst$ whenever needed.
   This means that a system that contains rule $A\isa B$ can never be brought into a state in which $\id{sat}_{A\isa B}$ is False.
   This allows for certain run-time optimizations, which are beyond the scope of this paper.
   A second property that sets specialization rules apart, is that they play a role in the Ampersand compiler.
   For example, code that determines equality between atoms only works for atoms that are an instance of the same concept.
   To ensure that this does not result in problems at run-time, users are required to write all specialization rules explicitly.

@@

   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user or an automated actor without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

   Every rule is assigned to at least one role by means of the relation $\maintain$.
   $o\ \maintain\ u$ means that role $o$ must keep rule $u$ satisfied, i.e.\ to keep $\sat{u}{\dataset}$ true.
   After some event has violated a rule $u$ (by making changes in the dataset),
   role $o$ maintains this rule by making other changes to satisfy rule $u$ again.
   We call this ``to restore the invariance'' of rule $u$.
   In other words: If anyone edits in the dataset and violates a rule, someone or something has to fix the dataset to satisfy that rule again.
   Example: if a rule says that every request must be acknowledged, the receipt of a request will violate this rule.
   Sending a message that acknowledges this receipt will cause the rule to be satisfied again.

   The fact that every rule has at least one role in the relation $\maintain$, means that all roles together are keeping all rules satisfied.

\subsection{Example}
\label{old IS}
   Having defined an information system in mathematical terms, let us give an example.
   For this purpose we use the language Ampersand
   because it makes the examples more appealing to read.
   Let us first define a dataset of seven triples and three relations.
\begin{verbatim}
RELATION takes[Student*Course] =
[ ("Peter", "Management")
; ("Susan", "Business IT")
; ("John", "Business IT")
]
\end{verbatim}
   This introduces a relation with name \verb#takes#,
   source concept \verb#Student#, and
   target concept \verb#Course#.
   It also introduces three triples:
\[\begin{array}{l}
   \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
   \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
   \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
\end{array}\]
   The informal meaning of this relation is that it states which students are taking which courses.
   So, there is a student, Peter, taking the course Management and two students, Susan and John, taking the course Business IT.

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}
   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that student takes.
   Its constraint, $p_{\tt Enroll}$ is written in logic as:
\[\begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\]
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to maintain it:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}

   Summarizing, this example contains a dataset with seven triples
   and a schema with three concepts, three relations, and one rule.
   Now let us define an information system.
   Notice that the current population satisfies the rule.
   Let us check requirements~\ref{eqn:inst} thru~\ref{eqn:specialization} to verify that this is an information system.
   Requirement~\ref{eqn:concepts} defines the set $\concepts$.
   Requirement~\ref{eqn:define R} defines the set $\rels$.
   Requirement~\ref{eqn:type} defines the relation $\id{inst}$.

%    The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
%    This means that every atom is an instance of a concept:
% \begin{eqnarray}
%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:inst}
% \end{eqnarray}
%    To do type checking based on the schema only,
%    all relations in triples from the dataset must be ``known'' in the schema:
% \begin{equation}
%    \forall a,r,b:\ \triple{a}{r}{b}\in\dataset\ \Rightarrow\ r\in\rels\label{eqn:define R}
% \end{equation}

%    In a typed information system, atoms in a relation have a type as declared in a relation.
%    For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
%    We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
% \begin{eqnarray}
%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:concepts}\\
%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\dataset&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
% \end{eqnarray}
%    The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.

%    To compare atoms for equality,
%    the set $\rels$ contains an identity relation $\ident{C}$ for every concept $C$ in which all atoms occur.
%    It satisfies:
% \begin{equation}
%    \forall a,b:\ a\ \inst\ C\ \wedge\ b\ \inst\ C\ \Rightarrow\ (a\ \ident{C}\ b\ \Leftrightarrow\ a=b)
% \end{equation}
%    So, $\id{pop}_{\ident{C}}$ is defined by:
% \begin{eqnarray}
%    \forall C\in\concepts&:&\pop{\ident{C}}{\dataset}=\{\pair{a}{a}\mid a\ \inst\ C\}
% \end{eqnarray}

%    Having one rule, one role, a relation $\maintain$ that contains one pair, and a dataset,
%    we have now defined a small system.

@@

   We use two datasets: $\dataset$ and $\dataset'$.
   Before doing so, let us first define the disjoint union of two systems.
\begin{definition}[disjoint union of datasets]
\begin{eqnarray}
   \omit\rlap{$\la\atoms,\concepts,\inst,\isa,\rels,\dataset\ra\sqcup\la\atoms',\concepts',\inst',\isa',\rels',\dataset'\ra$}\notag\\
   &=&\la\atoms\uplus\atoms',\ \concepts\uplus\concepts',\ \inst\uplus^2\inst',\ \isa\uplus^2\isa',\ \rels\uplus^3\rels',\ \dataset\uplus^5\dataset'\ra\notag\\
      X\uplus Y&=&\{(x,0)\mid\ x\in X\}\ \cup\ \{(y,1)\mid\ y\in Y\}\\
      X\uplus^2 Y&=&\begin{array}[t]{@{}l}\{((x_1,0),(x_2,0))\mid\ (x_1,x_2)\in X\}\ \cup\\ \{((y_1,1),(y_2,1))\mid\ (y_1,y_2)\in Y\}\end{array}\\
      X\uplus^3 Y&=&\begin{array}[t]{@{}l}\{\declare{(n,0)}{(A,0)}{(B,0)}\mid\ \declare{n}{A}{B}\in X\}\ \cup\\ \{\declare{(n,1)}{(A,1)}{(B,1)}\mid\ \declare{n}{A}{B}\in Y\}\end{array}\\
      X\uplus^5 Y&=&\begin{array}[t]{@{}l}\{\triple{(a,0)}{\declare{(n,0)}{(A,0)}{(B,0)}}{(b,0)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in X\}\ \cup\\ \{\triple{(a,1)}{\declare{(n,1)}{(A,1)}{(B,1)}}{(b,1)}\mid\ \triple{a}{\declare{n}{A}{B}}{b}\in Y\}\end{array}
\end{eqnarray}
\end{definition}
Note that if $\dataset$ and $\dataset'$ are datasets, then so is $\dataset\sqcup\dataset'$.

@@

\begin{definition}[migration dataset]
   \begin{eqnarray}
      &&\begin{array}[t]{@{}l}
         \{ {\tt ENFORCE\ }\declare{(nm',1)}{(A',1)}{(B',1)}\\\quad{\tt\ :=\ }\ident{(A',1)};\declare{(nm,0)}{(A,0)}{(B,0)};\ident{(B',1)}\\
            \mid\ \begin{array}[t]{@{}l}
               \declare{nm}{A}{B}\in\rels\wedge\declare{nm'}{A'}{B'}\in\rels'\wedge\id{nm}=\id{nm'}\ \wedge\\
               \{\pair{A}{A'},\pair{B}{B'}\}\subseteq\isa\cup\isa'\cup\flip{\isa}\cup\flip{\isa'}\}
               \end{array}
           \end{array}\\
           &\cup&\begin{array}[t]{@{}l}
            \{ {\tt CLASSIFY\ }(C',1){\tt\ IS\ }(C,0)\mid\ C\in\concepts,C'\in\concepts',C=C'\}\\
           \end{array}
   \end{eqnarray}
\end{definition}

@@

\subsection{Changes}
% The purpose of this section is to explain why a dataset is structured the way it is.
   In the migration of a dataset we deal with changes to the elements that are not data:
   $\inst$, $\concepts$, and $\rels$.
   Such changes have further reaching consequences, however.
   Changes to $\inst$, $\concepts$, $\rels$ change the data structure in a dataset.
   We define a \define{migration of a dataset} $\la\atoms,\concepts,\inst,\rels,\dataset\ra$ as a change in which one or more of $\inst$, $\concepts$, or $\rels$ change.
   To satisfy the definition of datasets, $\dataset$ and $\atoms$ must change too,
   but the migration should try to preserve the maximal amount of data.
